---
title: "Getting Started in R and R Studio"
output:
  html_document:
    theme: cerulean 
    highlight: tango
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Right now, you're working inside of RStudio in an R markdown document! 

R markdown is great because it allows you to:

- Combine code and text 
- Display both the output of the code as well as the code itself
- Knit to an .html document to easily share your work with colleagues
  - Knit is kind of a funny word to use here — it comes from knitting! The idea is that you're combining text, code, and output together seamlessly, like knitting pieces of yarn into a sweater. In fact, the Knit button even has a little ball of blue yarn and a knitting needle!

We're going to knit to an .html in this course, but you can also output to a PDF or word document. 

Inside of the R markdown document, you can type out text. That is what this is! You can include code chunks, also.

This is an example of an empty code chunk:

```{r}
```

Inside of code chunks, you can include both **code** and **comments**, which are any text after the `#` symbol. Descriptive comments are super important! They help you remember what your code is doing if you access it later, and also make it clear for other people who view your code.

This code chunk has a comment in it:

```{r}
### This is a test comment
```

Now, let's include some code. To start - R can be used as a calculator, using the standard arithmetic symbols. 

To execute this code chunk, you should see a green triangle (a "play" symbol) in the upper-right hand corner of the code chunk. When you press play, hopefully you see 6 output below the code chunk!

```{r}
### Add 3 and 3 together
3 + 3
```
Here are some other numeric operations:

```{r}
### Division 
12 / 3
### Subtraction
12 - 3
### Multiplication 
12 * 3
### Anything in the parentheses is executed first
(3 + 3) ^ 2 ### The ^ symbol is used for exponents
```
## Try it!

1. In this code chunk, return the sum of two and eight.

Please keep free to copy and paste code from above! If you do so, just copy the code, not the back ticks starting and ending the code chunk, or the {r}.

```{r}

```

2. In Miguel's progress report from X, he reported that there were X participants enrolled in Eswatini, Y in Tanzania, and Z in Eswatini. How many total participants were enrolled?

```{r}

```

3. It turns out that Y participants from Eswatini are missing z. We started with z participants, so how many have available data for z?

```{r}

```

# Basic R skills

Using R as a calculator is pretty simple - and just scratching the surface of what R is capable of! 

R can be used for analysis, including t-tests, chi-squared tests, ANOVAs, linear regression, and logistic regression. It can also be used for complicated analysis like machine learning. And, you can also make publication-quality figures in R!

R can also be used to do a lot of the same executions as Excel, such as sorting, filtering, and creating new variables. However, R is reproducible - you don't need to remember exactly what columns you selected or what operation you did because you'll have code to recreate any of these steps that you took. R handles big data sets better, and is more efficient for complicated calculations. 

One cool thing about R is that it is R and open source. This is in contrast to other tools - like SAS, Stata, and Prism which are not free. 

We will cover basic analysis and figures in the weeks leading up to the workshop and during the workshop itself. However, today we're going to get familiar with some basic R syntax and data structures. 

First, `<-` is used as an assignment operator. In the code chunk below, `var_a` is the assigned the value of 3 and `var_b` is assigned the value of 3. 

```{r}
var_a <- 3
var_b <- 3
```

You can do the same operations on `var_a` and `var_b` that we did above and get the same result:

```{r}
var_a + var_b 
```

Here, we used variable names for numbers. However, we can assign things like data frames and characters the same way! 

Up until now, we've been working with numeric values. However, there are many different data types in R. 

| Type       | Description                  | Example         |
|------------|------------------------------|-----------------|
| numeric    | Numbers         | `3.14`, `-2`    |
| character  | Text or strings              | `"hello"`       |
| logical    | Boolean values               | `TRUE`,


# Getting help

There are a few ways to get help in R.

First, you can include a `?` in front of a function name. For example, if you want more information on the `str()` function, the documentation should appear in the help() pane once you run the code chunk. This is helpful to see all of the function arguments. 

```{r}
?str()
```

I also Google - a lot! If you're stuck on something, chances are someone else has experienced the same error. Make sure to include the function name and the error you're experiencing. I usually also include "R" in my search to make sure to get results for the function in R, not another coding language.  

# Survey

Please fill out this survey so we know you completed the assignment and to help guide the direction this workshop is going in!

# TBeyond Basics

There are style guidelines for coding related to naming variables and syntax. Read this if you are interested in best practices!

https://style.tidyverse.org/


