---
title: "Manipulating data"
output:
  html_document:
    theme: cerulean 
    highlight: tango
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Manipulating data

This week, we're going to start with manipulating data.

There are a few different ways to do this. One is using the built-in R functions (called "base" R). Another is using the `dplyr` package.

We're going to focus primarily on `dplyr`. I find this to be more intuitive and quicker to learn than the base R equivalent. However, I'm also including a section at the bottom with resources on base R functions in case that may be helpful.

## Background information on dplyr

Please take 20 minutes to use these resources to learn about what exactly `dplyr` is and examples of some of the functions we're going to learn to better orient yourself before continuing.

https://dplyr.tidyverse.org/articles/dplyr.html

This is another great cheat sheet that includes lots of `dplyr` functions:

https://posit.co/wp-content/uploads/2022/10/data-transformation.pdf

Here is a good video that walks you through some of the functions you read about in the first link:

https://www.youtube.com/watch?v=Gvhkp-Yw65U&ab_channel=Dataslice

**Note**: we are still going to use `library(tidyverse)` to read in the `dplyr` package. This is because tidyverse is a **bundle** of some really useful packages - including `dplyr` which we'll use here and `ggplot2`, which we'll use in the next assignment. 

## Filtering

The first operation we are going to learn is how to filter to different rows in the data.

Often, we're interested in a specific time point or a specific subgroup of participants.

To start, read in the data. Please reference week 2 if you need a reminder.

```{r}
### Update the file path here to the location that you saved the data

### Right now, this is my file path - so the original code will give you an error if you try and run it

### Please review the week 2 video if you are stuck on this step

data <- read.csv("C:/Users/u249135/OneDrive - Baylor College of Medicine/Teaching Materials/R-workshop/PreWorkshop/Week3/Stool4TB Database for Abi.csv")
```

This dataset includes the same variables as week 2, though it includes all 3 sites, noted by the `rf_coutnry` variable.

It's always a good idea to start by taking a look at the data. The `head()` function allows us to take a look at the first few rows of the data

```{r}
head(data)
```

In week 2, I modified the data in order to send a clean version. However, what we're looking at now is the data extracted from RedCap.

We can see that each participant has more than one row in the data. Each row corresponds to a different `redcap_event_name` value. For example, `record_id` 1 has 4 rows in the data. Each row corresponds to a different time point, as well as a medication log.

Here is the code to filter to just the baseline data.

Often, we want to access the subset of data later. Let's save this subset as `baseline_data` - hopefully, you see it in the "Data" tab in your environment. 


```{r warning = FALSE, message = FALSE}
library(tidyverse)  ### You should have installed this package in week 2
baseline_data <- data%>%
  filter(redcap_event_name == "Baseline")
```

What’s happening here?

We briefly introduced the pipe operator `(%>%)` last week . Let’s walk through what this code is doing:

1. Start with data (the full dataset)

2. The pipe operator `%>%` means “and then” - it passes the result (in our case, the full dataset) from one step to the next.

3. `filter()` is a function from the dplyr package that keeps only the rows that meet a specific condition. In this case, it keeps only the rows where the value of redcap_event_name is "Baseline".

The final result is a version of your dataset that includes only the baseline records.

Sometimes, we may want to filter based on a few different variables. 

We can filter to all people with HIV from Eswatini, and save as a subset using the `&` operator.

```{r}
eswatini_pwh <- data%>%
  filter(rf_country == "Eswatini" & bf_hivstatus == "Positive")
```

If you prefer, we could also use `filter()` twice - this will give the same output.

```{r}
eswatini_pwh <- data%>%
  filter(rf_country == "Eswatini")%>%
  filter(bf_hivstatus == "Positive")
```

As a quick check, we can verify that we filtered to just the rows with "Eswatini" for `rf_country` and "Positive" for `bf_hivstatus`.

```{r}
table(eswatini_pwh$rf_country)%>%
  addmargins()
table(eswatini_pwh$bf_hivstatus)%>%
  addmargins()
```

We can filter on numeric columns, in addition to categorical variables. Suppose we want to filter to all participants who are 100 kg or heavier at baseline.

```{r }
data%>%
  filter(redcap_event_name == "Baseline")%>%
  filter(bf_weight >= 100)%>%
  head()
```

Filtering is also very helpful in identifying participants with missing data.

Do you remember how last week we identified a few participants who were missing the `rf_age_calculated` variable?

Let's filter to all participants at baseline who are missing this value to investigate this more. 

```{r}
data%>%
  filter(redcap_event_name == "Baseline" & is.na(rf_age_calculated))
```

If we take a look at the bottom of the output, we see that there are 1,432 rows missing this variable! 2 are from Eswatini - which are the participants we saw last week - and the rest are from Mozambique and Uganda. 

Pull up `data` in a new tab by clicking on `data` in your environment and take a look at this variable. It seems that this variable is just for participants from Eswatini. However, we can create a new variable for age by calculating this using the two date columns - `rf_enroll_date` and `rf_dob`. 
 
## Creating new variables

The `mutate()` function from `dplyr` allows us to create new variables.

Once again, we are going to use the pipe operator (`%>%`).

Creating a new variable using the two date fields is going to be a little complicated, so let's start with something simpler.

To start, let's create a column, `study`, that is **Stool4TB** for all participants. In order for the changes to be saved in `data`, we need to make sure to assign this as `data`.

```{r}
data <- data%>%
  mutate(study = "Stool4TB")
```

Some important things to note:

1. The `filter()` function used two equals signs, however `mutate()` just uses one. 

  - In `mutate()`, you are creating or modifying a column. You use one equals sign (=) to assign values to a new or existing variable. 

  - In `filter()`, you're testing for a condition. To check whether a variable is equal to a value, you use two equals signs (==).

2. It's important that we put the study name - "Stool4TB" in quotation marks. If we don't, we will see this error:

```{r eval = FALSE }
data <- data%>%
  mutate(study = Stool4TB)
```

Let's take a look at this variable!

```{r}
table(data$study, exclude = NULL)%>%
  addmargins()
```

We've just created a new variable called `study` that has "Stool4TB" for each row in the data.

This is a bit simple - let's move on to creating a variable for age. 

Right now, the structure of each date is a character. 

```{r}
str(data$rf_enroll_date)
str(data$rf_dob)
```

However, we can convert each to a **date** type, and then subtract the date of birth from the enrollment date to get the age they were at enrollment.

Let's break this down in steps. First, let's convert each column to a **date** type. To do this, we're going to use the `ymd()` function in the `lubridate` package.

`ymd()` tells R that the date is formatted in year, month, day order. However, `lubridate` has other functions if you are working with dates that are formatted differently, such as `mdy()`.

```{r}
### Remember, you only need to install packages once
### install.packages(lubridate)
library(lubridate)
data <- data%>%
  mutate(rf_enroll_date = ymd(rf_enroll_date))%>%
  ### You can string together multiple functions using the pipe operator!
  mutate(rf_dob = ymd(rf_dob)) 

### Notice here that I am saving each variable as the original variable name 
### However, you could also create a copy of each variable by naming it something else if you prefer
```

```{r}
str(data$rf_enroll_date)
str(data$rf_dob)
```

Hopefully you see that both variables are now `Date` type. However, you will also see some `NA` values! This could be because the dates are only collected at enrollment - the "Baseline" `redcap_event_name` value. However, this could also be because some dates are missing at baseline. We will investigate this shortly.

Next, let's find the difference between the two dates.

```{r}
data <- data%>%
  ### Create a new variable, date_diff
  
  ### interval and time_length are both functions in the lubridate package
  
  ### interval creates an interval between two dates - this will give something like Interval: 2018-01-01 to 2020-06-01
  
  ### time_length finds the length of time in the interval, and by specifying unit = "years" we get the difference in years versus days or months 
  mutate(date_diff = time_length(interval(rf_dob, rf_enroll_date), unit = "years"))
```

We're getting close! However, upon inspecting the variable, it seems like  `date_diff` is reporting years out to a decimal place. We want to know their integer age, rounding down to the nearest whole number.

The `floor()` function rounds down to the nearest integer.

```{r}
data <- data%>%
  mutate(date_diff = floor(date_diff))
### Hint - if you ever need to round to the nearest integer or the nearest decimal place, you can use the round() function
```

We're moving in the right direction!

```{r}
table(data$date_diff)
```

## Renaming variables

We just created a variable called `date_diff`. However, there is probably a better, more informative name for this.

Let's rename this variable to be `calculated_age`.

Again - we're going to use the same pipe operator with a new function - `rename()`. The new name goes first, followed by the old name.

```{r }
data <- data%>%
  rename(calculated_age = date_diff)
```

## Selecting columns

At this point, let's confirm that everyone at baseline has an age reported.

We want to filter to all participants at baseline who are missing this newly created variable, and select just the important columns related to participant age.

This is where we'll use a new function - `select()`.

```{r}
data%>%
  ### Filter to all participants at baseline missing calculated_age
  filter(redcap_event_name == "Baseline" & is.na(calculated_age))%>%
  select(record_id, calculated_age, rf_dob, rf_age, rf_enroll_date)
```

143 participants do not have this value. And, it looks like this is missing because they are missing `rf_dob`.

However, they do have another variable - `rf_age`. Let's use the `rf_age` variable if the calculated age is missing. To do this, we're going to use the `ifelse()` function. Essentially, if the first condition is met, go with the first value. Else, go with the second value.

So here - if the `calculated_age` value is NA, use the value in `rf_age`. Else - if the `calculated_age` value is not missing, retain this age.

Let's save as a new variable - `composite_age` - so we know that this is a combination of our calculated age and the `rf_age` variable. 

```{r}
data <- data%>%
  mutate(composite_age = ifelse(is.na(calculated_age), rf_age, calculated_age))
```

As a final check, is anyone still missing age at enrollment?

```{r}
data%>%
  filter(redcap_event_name == "Baseline" & is.na(composite_age))%>%
  select(record_id, composite_age, rf_dob, rf_enroll_date, rf_age)

```

One participant is missing `rf_dob` so we couldn't calculate the age, and the `rf_age` value is also missing

This should be a query to the clinical teams so they can track down this information.

## Arrange

Lastly, `dplyr` has a function, `arrange()` that can order numeric variables.

For example, let's filter to the baseline data, and then sort by the `bf_weight` variable.

```{r}
data%>%
  filter(redcap_event_name == "Baseline")%>%
  arrange(bf_weight)%>%
  select(record_id, bf_weight)%>%
  ### Rather than printing out all the rows, let's use the head() function to take a peek at the first few rows 
  head()
```

This sorts the data in ascending order for weight. We can easily sort in descending order by including `desc()` in the code:

```{r}
data%>%
  filter(redcap_event_name == "Baseline")%>%
  arrange(desc(bf_weight))%>%
  select(record_id, bf_weight)
```

Now, on to some practice! 

# Correct my mistakes

## Fix it! 1

I want to filter to all participants from Eswatini at all time points. However, I'm getting an error! My code returns a data frame with 0 rows.

```{r eval = FALSE}
data%>%
  filter(rf_country == "Eswatini ")
```



```{r}
### Hint: take a look at what the `rf_country` column looks like. I often use the `table()` function for this. Refer back to the week 2 assignment

```

## Fix it! 2

I want to filter to all participants at baseline, and save as a subset called `baseline_subset`. However, I'm getting an error - my code doesn't run.

```{r eval = FALSE}
baseline_subset <- data%>%
  filter(redcap_event_name = "Baseline")
```

## Fix it! 3

I want to filter to the missing data in the `composite_age` column. This is returning 0 rows, however we already identified 1 participant with missing data. Something is wrong - can you fix it? 

```{r eval = FALSE}
data%>%
  filter(redcap_event_name == "Baseline" & composite_age == "")
```

### Fix it! 4

I want to rename the `bf_weight` column to be `baseline_weight`. However, I am unable to get the code to run.

```{r eval = FALSE}
data%>%
  rename(bf_weight = baseline_weight)
```

## Fix it! 5

I want to create a new column called `60_weight` that is "Weight is less than 60kg" if `bf_weight` is less than 60 and "Weight is greater than or equal to 60kg" if not.

Update the code so this works.

**Hint**: we haven't reviewed this specific error before. One important thing to learn when coding is how to search for the solution to coding error! Please Google or use ChatGPT to figure out why this is causing an error and how to fix it.

```{r eval = FALSE}
data%>%
  filter(redcap_event_name == "Baseline")%>%
  mutate(60_weight = ifelse(bf_weight < 60, "Weight is less than 60kg", "Weight is greater than or equal to 60kg"))
```

# Practice

If you get stuck on any of the "Try it!" questions, please review the code and notes above. You've learned all the operations you need to complete these questions. 

## Try it! 1

Filter to all participants from Mozambique who are female. Select the column for participant ID and respiratory rate, and arrange the respiratory rate column in descending order. 

```{r}

```

**Above and beyond** - can you figure out how to print out just the 10 highest respiratory rates?

## Try it! 2

After consulting with some clinicians, they tell you that any heart rate above 150 beats/minute may be extreme and to provide a list of all participants who meet this condition in order to double-check that the data is correct.

Filter to all heart rates above 150 beats/minute at baseline, and select the heart rate and participant ID columns. 

Save the result as `heartrate_investigation_subset`.

```{r }

```

**Above and beyond** - in most cases, team members who are resolving queries require an Excel spreadsheet. We have learned the `read.csv()` function. There is another function `write.csv()` that exports the data to a .csv file.

If you did Try it! 2 above correctly, the code chunk below should save the output to a .csv file in the same file location that your .Rmd and data are saved to. 

```{r eval = FALSE }
write.csv(heartrate_investigation_subset, "Example Export.csv")
```

## Try it! 3

In week 2, you learned how to perform summary statistics on the data.

Here, I want you to create two subsets. Save each as a separate data frame - you can choose what to name them.

The first should include children < 8 at baseline who are living with HIV.

The second should include children < 8 at baseline who are not living with HIV.

**Please remember to use the composite_age variable that we created above!**

```{r}
### Create subset 1 

```

```{r}
### Create subset 2

```

I want you to compare 3 different summary statistics between these two subsets. You may choose to use any of the variables in the data to summarize. You could create some tables and report the proportions in each group, or investigate some of the numeric variables.

**Please include na.rm = TRUE in any numeric summary functions you choose to use! There are a small number of missing values, and the summary statistic will be NA if you do not include this.**

```{r}
### Summary statistics for subset 1
```

```{r}
### Summary statistics for subset 2
```

Provide commentary on how these two groups are the same or different. 

# Base R

There are different ways to manipulate data. What you've learned so far has been using the `dplyr` package. I find `dplyr` to be more intuitive and quicker to learn than base R. However, there are certain situations where base R may be helpful to know. Here is a resource to learn more about base R data manipulation if you are interested!

https://stats.oarc.ucla.edu/r/modules/subsetting-data/



