---
title: "Workshop part 2"
output:
  html_document:
    theme: cerulean 
    highlight: tango
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For the second half of the workshop, we're going to be working on a bigger piece of analysis. The goal is to turn this in to a manuscript.

# Research question

Working with Anna, Anca, and Joanna, the following research question was developed:

What are the clinical traits associated with having a stool Xpert Ultra trace result?

# Set-up

Start by reading in the data. This is the same data that we've been using for the pre-workshop activities, with a few additional columns. 

```{r}
data <- read.csv("C:/Users/u249135/OneDrive - Baylor College of Medicine/Teaching Materials/R-workshop/Workshop/Stool4TB Database for Abi.csv")
```

```{r}
### Let's start by taking a look at the column names of the data
colnames(data)
```

We're going to be using a few packages that we've learned today, so read in the `dplyr`, `ggplot2`, and `lubridate` packages. 

```{r warning = FALSE, message = FALSE }
library(tidyverse)
library(lubridate)
```

We will also use a few new packages today. Please install a few additional packages. 

```{r}
install.packages("GGally")
```

```{r}
library(GGally)
```

For the analysis question, we are interested in the stool Xpert Ultra trace results. So to start, let's filter to all participants who have a positive stool Xpert Ultra result. 

```{r}
data <- data%>%
  ### Filter to all participants at baseline
  filter(redcap_event_name == "Baseline")%>%
  ### Filter to those with a positive stool Xpert Ultra result
  filter(pe_stoolxpertresult == "MTB Detected")
```

Let's take a peek at the semi-quantitative results of the stool Xpert Ultra test:

```{r}
table(data$pe_stoolsemiresult, exclude = NULL)%>%
  addmargins()
```

Create a variable with whether or not a participant has a stool Xpert Ultra trace result.

```{r}
data <- data%>%
  ### Create a variable that is "Trace" if the semi-quantitative value is "Trace" and "Not Trace" otherwise
  mutate(stool_semiquant_bin = ifelse(pe_stoolsemiresult == "Trace", "Trace", "Not Trace"))
```

# Where to start

We are going to build a **logistic regression model**.

Logistic regression models are used to predict an outcome with **2 levels**. In our case, we want to predict whether a stool Xpert Ultra will be trace or not. 

We are going to use other variables in the data to make an informed prediction.

# Create a few variables

```{r}
### BMI is not a variable in the data, but weight and height are
data <- data%>%
  ### Create a variable for BMI
  
  ### Height is in centimeters, so divide by 100 to convert to meters 
  mutate(bmi = bf_weight / ((bf_height / 100)^2))
```

```{r}
### Create a variable for age 
### This should be familiar - we did this in the dplyr assignment

data <- data%>%
  ### Convert the dates to a date data type
  mutate(rf_enroll_date = ymd(rf_enroll_date))%>%
  mutate(rf_dob = ymd(rf_dob))%>%
  ### Find the difference, in years, between the dates
  mutate(date_diff = time_length(interval(rf_dob, rf_enroll_date), unit = "years"))%>%
  ### Round down to the nearest integer age
  mutate(date_diff = floor(date_diff))%>%
  ### Some dates were not able to be calculated due to missing dates
  mutate(age = ifelse(is.na(date_diff), rf_age, date_diff))
```

```{r}
## Create a variable for CD4 result
data <- data%>%
  ### Convert both variables to be numeric
  mutate(bf_cd4result = as.numeric(bf_cd4result))%>%
  mutate(bf_recentcd4 = as.numeric(bf_recentcd4))%>%
  ### If the result at enrollment is missing, use the recent CD4 result
  mutate(composite_cd4 = ifelse(is.na(bf_cd4result), bf_recentcd4, bf_cd4result))
```

# What predictors to include?

We've defined the research question. But, how do we know what variables to include that may explain whether a participant has a trace or not trace stool Xpert Ultra result?

There are different ways to approach this.

1. If we were to take more of a machine learning approach, we could include **all** of the variables that we collected as predictors in the model.

2. We could include a penalty for each variable included in the model by fitting a penalized regression model, like lasso or ridge regression.

3. We could fit a logistic regression model with *all* the predictors and then perform backwards variable selection to find an optimal subset of variables.

4. We could fit a logistic regression model with a small subset of variables that subject matter experts have deemed important.

How do we choose what approach to use? 

For this analysis, we're going to conduct this with the subset of variables that clinicians have decided are important.

* Age
* Country
* Sex
* Enrollment location
* BMI
* CD4 count, in participants living with HIV 

## Think about it!

For the workshop today, we are just going to use the variables listed above.

However, what are some other variables that may be related to a participant having a stool Xpert Ultra trace result?

# Visualize the relationship of the variables to each other

When doing an analysis like this, the first - and a very important - step is to look at the data. By creating tables and figures, we can investigate:

1. Is there any missing data?

2. Are there any outliers?

3. What are any trends or relationships in the data? Are any of the predictors strongly correlated with each other?

This will inform us both if there is any data cleaning that still needs to be done, as well as the variables we ultimately include in the model.

To start, consider the relationship between the predictors. Consider `bf_weight`, `bf_height`, and our newly created variable `bmi`.

We can use the `ggpairs()` function form the `GGally` package to get useful information about the relationship between multiple variables without having to create a ton of figures. This function provides a scatterplot and correlations for each pair as well as density plots for each variable.

## Height, weight, and BMI

```{r}
### Inside the ggpairs function, include a subset of the variables that we want to include in the figure
ggpairs(data%>%
          select(bf_weight, bf_height, bmi))
```

## Think about it!

What relationships do you see in the data? There's a lot of information here - can you figure out what is being plotted?



Multicollinearity (when two or more variables in a model are highly correlated) can lead to several problems in model estimation:

1. **Unstable model estimates**: when predictors are highly correlated, small changes in the data can cause large swings in the estimated coefficients.
2. **Difficult interpretation**: when predictors convey overlapping information, it becomes challenging to isolate the effect of each one on the outcome.
3. **Reduced predictive power**: While the model may fit the training data well, multicollinearity can hurt generalization to new data.

## Age and HIV status

Let's look at the relationship between age and HIV status. First, we need to modify the variable for HIV status, because it was just collected for the 0-8 subgroup and is missing for the adults. 

```{r}
### We did this in the first portion of the workshop activity

data <- data%>%
  ### Create a new variable, "new_hiv_status"
  
  ### When the participant is in the 15 and older category AND the HIV status is missing, the value is positive
  
  mutate(new_hiv_status = case_when(age_cat == "15 and older" & bf_hivstatus == "" ~ "Positive",
  ### When that condition is NOT met (so for all the participants in the 0-8 category), retain the value of the bf_hivstatus variable
                                TRUE ~ bf_hivstatus))
```

Let's investigate the relationship between age and HIV status:

```{r}
table(data$age_cat, data$new_hiv_status)%>%
  addmargins()
```

100% of adults are living with HIV. This means there's no variation in HIV status among adults - their HIV status is completely determined by age. This is known as **perfect separation** or between age and HIV status.

This is a problem for logistic regression modeling.

In logistic regression, each predictor should explain unique, independent information about the outcome. When one predictor (like age) perfectly predicts another (like HIV status), the model can't tell which one is truly driving the relationship with the outcome. This can lead to:

* Unstable or infinite estimates
* Large standard errors
* Model convergence issues
* Misleading results

Currently, it doesn't make sense to include both age and HIV status in the model with all participants. 

# What subset to use moving forward? 

In light of what we've investigated so far, we have a few options. 

## 1. Just adults.

This would allow us to include CD4 in the model, because all adults are living with HIV.

```{r}
adult_subset <- data%>%
  filter(age_cat == "15 and older")
```

```{r}
sum(is.na(adult_subset$composite_cd4))
nrow(adult_subset)
```

```{r}
57 / 59 
```

This variable is 96.6% complete; just 2 adults living with HIV are missing CD4 count. 

## 2. Just children

This would allow us to include HIV status in the model.

```{r}
children_subset <- data%>%
  filter(age_cat == "0-8")
table(children_subset$outcome, children_subset$bf_hivstatus, exclude = NULL)%>%
  addmargins()
```

However, only 2 participants are living with HIV, and both have a stool Xpert Ultra trace result. This could potentially lead to model fit issues.

## 3. Include both age categories

If we include all ages, we won't be able to include HIV status because of the perfect separation we identified above. This is a limitation, because it will be difficult to tease out if the effect is due to HIV status or age. 

## Think about it!

Moving forward in this workshop, we will just use the adult analytic cohort and include the CD4 count as a variable in the model. 

However, what should we do differently for the manuscript?

```{r}
### Before proceeding, remove the 2 participants missing CD4 from the analysis
adult_subset <- adult_subset%>%
  filter(!is.na(composite_cd4))%>%
  ### Create a binary variable from CD4 
  mutate(cd4_cat = ifelse(composite_cd4 < 200, "CD4 < 200", "CD4 >= 200"))

### There are other approaches for handling missing data, including impution. However, today we'll proceed with dropping the 2 participants. 
```

```{r}
### How many participants does this include?
nrow(adult_subset)
### What does our outcome - the newly created variable - look like?
table(adult_subset$stool_semiquant_bin, exclude = NULL)%>%
  addmargins()
```

There are 57 participants in the adult_subset. Of these, 32 have a trace result. 

```{r}
32 / 57
```

# Visualize the variables in relation to the outcome

Now, let's investigate the relationship between this outcome and the variables we're going to include in the model.

```{r}
### Let's take a look at this variable
ggplot(adult_subset = adult_subset, aes(x = bmi))+
  geom_histogram()+
  labs(x = "BMI",
       y = "Count",
       title = "Histogram of BMI",
       subtitle = "Among participants with a positive stool Xpert Ultra result at baseline")
```

For each variable, plot the variable along with the outcome.

## Country

```{r}
ggplot(adult_subset = adult_subset, aes(x = rf_country, fill = stool_semiquant_bin))+
  geom_bar()+
  labs(x = "Country",
       y = "Count",
       fill = "Stool Xpert Ultra Semi-Quantitative Bin",
       title = "Stacked Barplot of Stool Xpert Ultra Semi-Quantitative bin by Country",
       subtitle = "Among participants with a positive stool Xpert Ultra result at baseline")
```

## BMI

```{r}
ggplot(adult_subset = adult_subset, aes(x = stool_semiquant_bin, y = bmi))+
  geom_boxplot()+
  labs(x = "Stool Xpert Ultra Semi-quantitative Bin",
       y = "BMI",
       title = "Boxplots of BMI by stool Xpert Ultra semi-quantitative bin",
       subtitle = "Among participants with a positive stool Xpert Ultra result at baseline")
```

## Enrollment location

```{r}
table(adult_subset$rf_enrollocation, adult_subset$stool_semiquant_bin, exclude = NULL)%>%
  addmargins()
```

```{r}
ggplot(adult_subset = adult_subset, aes(x = rf_enrollocation, fill = stool_semiquant_bin))+
  geom_bar()+
  labs(x = "Enrollment location",
       y = "Count",
       fill = "Stool Xpert Ultra Semi-Quantitative Bin",
       title = "Stacked Barplot of Stool Xpert Ultra Semi-Quantitative bin by Enrollment location",
       subtitle = "Among participants with a positive stool Xpert Ultra result at baseline")
```
We can see that there are just **3** participants who were enrolled in an inpatient setting and the rest enrolled in an outpatient setting. We won't include this in the final model, because it will create some problems:

**Very small group size**: With just 3 inpatient participants, the model has very little adult_subset to reliably estimate the effect of being inpatient.

**Unstable estimates**: The regression coefficient for inpatient will be highly unstable and imprecise, leading to wide confidence intervals or even convergence problems.

**Low statistical power**: With so few cases, it’s unlikely we'll detect a meaningful association even if one exists.

**Potential overfitting**: Including rare categories can cause the model to fit noise rather than real patterns, hurting generalizability.

## Gender

```{r}
table(adult_subset$rf_gender, adult_subset$stool_semiquant_bin, exclude = NULL) %>%
  addmargins()
```

We're going to pause here to introduce a new concept: **odds ratios**.

## Odds ratios

An **odds ratio (OR)** is a way to compare the **likelihood of something happening in one group compared to another**.

In this case, we're comparing the likelihood of having a stool Xpert Ultra trace result between two groups - **females and males**.

### Step 1: What are “odds”?

The **odds** of an event are calculated as:

> odds = probability of the event happening ÷ probability of it not happening

Let’s calculate the odds for **females**:

- 35 had a trace result  
- 19 did not

```{r}
35 / 19
```

Now let’s calculate the odds for **males**:

- 22 had a trace result  
- 13 did not

```{r}
22 / 13
```

### Step 2: Calculate the Odds Ratio

The **odds ratio (OR)** compares the odds in females to the odds in males.

```{r}
(35 / 19) / (22 / 13)
```

### Step 3: Interpretation

The odds of having a stool Xpert Ultra trace result are approximately **1.089 times higher in females than in males**. Here, males would be the **reference category**.

An odds ratio of **1** means that both groups have the same likelihood of experiencing the event.

An odds ratio **greater than 1** means that the group being compared has higher odds of the event happening than the reference group. In other words, the event is more likely in this group.

An odds ratio **less than 1** means that the group being compared has lower odds of the event happening than the reference group. In other words, the event is less likely in this group.

### Step 4. Connecting this to logistic regression

We can do this same calculation by fitting a **logistic regression model with one variable**.

```{r}
adult_subset <- adult_subset%>%
  ### First, create a variable - outcome - that is 1 if Trace, 0 if not. This is necessary to fit the model.
  mutate(outcome = ifelse(stool_semiquant_bin == "Trace", 1, 0))%>%
  ### By default, R chooses the first alphabetical category to be the reference category. However, often we want the OR > 1 to make it easier to interpret. Because the have already shown that the OR > 1 when the reference is Male, convert to a factor and switch the order
  mutate(rf_gender = factor(rf_gender, levels = c("Male", "Female")))

### Build the model!
logistic_regression_model <- glm(outcome ~ rf_gender, adult_subset = adult_subset, family = "binomial")
```

```{r}
summary(logistic_regression_model)
```

There is a lot in this summary! We'll learn more about some of the other values later, but let's focus on the coefficient estimate for `rf_gender.`

We see `rf_genderFemale`. This let's us know that the odds are for the female group, so we know the **reference group** must be male.

The estimate is 0.08482. A logistic regression model predicts the **log odds** rather than the odds themselves. This is because odds are bounded between 0 and infinity, which can cause difficulties with modeling. By taking the **log**, we transform the odds to a scale that ranges from negative infinity to infinity, making it easier to model with a linear equation.

However, the odds ratio is easier to interpret than the log odds. 

To go from the **log odds** to the **odds ratio** we can exponentiate the coefficient estimate.

```{r}
exp(0.08482)
```

This is the same value we calculated above by hand! This is what is going on "behind the scenes" of the function, and the same logic applies to a logistic regression model with multiple predictors. 

Now, let's pivot back to visualizing the relationship between the other variables and the outcome. 

## Age 

```{r}
ggplot(adult_subset = adult_subset, aes(x = stool_semiquant_bin, y = age))+
  geom_boxplot()+
  labs(x = "Stool Xpert Ultra Semi-quantitative Bin",
       y = "Age (years)",
       title = "Boxplots of age by stool Xpert Ultra semi-quantitative bin",
       subtitle = "Among participants with a positive stool Xpert Ultra result at baseline")
```

We can represent numeric variables as categorical by creating bins. This is commonly done for variables such as age, when we know that younger children have a different response to TB than older adults.

```{r}
table(adult_subset$age)%>%
  addmargins()
```

Create a categorical variable from age, 0-8 and 15 and older.

```{r}
adult_subset <- adult_subset%>%
  mutate(age_cat = ifelse(age <= 8, "0-8", "15 and older"))
```

```{r}
table(adult_subset$age_cat, exclude = NULL)%>%
  addmargins()
```

Finally, let's look at the relationship between age category and the outcome.

```{r}
table(adult_subset$age_cat, adult_subset$stool_semiquant_bin, exclude = NULL)%>%
  addmargins()
```
```{r}
ggplot(adult_subset = adult_subset, aes(x = age_cat, fill = stool_semiquant_bin))+
  geom_bar()+
  labs(x = "Age category",
       y = "Count",
       fill = "Stool Xpert Ultra Semi-Quantitative Bin",
       title = "Stacked Barplot of Stool Xpert Ultra Semi-Quantitative bin by Age",
       subtitle = "Among participants with a positive stool Xpert Ultra result at baseline")
```

## Think about it! 

Discuss with someone near you - what does the above table and figure tell us about the relationship between age and whether a stool Xpert Ultra result is trace or not?

## Try it! 

```{r}

```

# Build the model

```{r}
table(adult_subset$cd4_cat, exclude = NULL)%>%
  addmargins()
```

```{r}
final_model <- glm(outcome ~ rf_country + rf_gender + bmi + cd4_cat, adult_subset = adult_subset, family = "binomial")
```

```{r }
summary(final_model)
```

# Model diagnostics 

The assumptions for logistics regression models are:

1. The outcome is binary
2. There are no extreme values or outliers
3. There are no highly correlated predictors
4. There is a linear relationship between the **log odds** of the outcome and each predictor value.

Let's tackle these one at a time!

## 1. The outcome is binary

We have ensured that with the construction of our model

## 2. There are no extreme values or outliers

```{r}
### install.packages("broom")
library(broom)

model.data <- augment(final_model) %>% 
  mutate(index = 1:n()) 

ggplot(model.data, aes(index, .std.resid)) + 
  geom_point(aes(color = as.factor(outcome)), alpha = .5) +
  theme_bw()+
  labs(x = "Generic index for each participant",
       y = "Standardized residual",
       color = "Outcome")
```


```{r}
data <- data %>%
  mutate(predicted_logit = predict(final_model, type = "link"))  # link = logit

# Plot logit vs. continuous predictor
ggplot(data, aes(x = bmi, y = predicted_logit)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Logit vs. BMI", y = "Logit (log-odds)", x = "BMI")
```

# Interpret the model



# Supplementary analysis 



## ATT initiation

## Outcome

# Discussion of results

1. What additional analysis should be included in this manuscript?

2. Are any of the results of the model surprising?

3. What are some of the limitations of our results?

4. What are of the main points of the analysis?

